A Generalized Sphere Theorem for Positively Curved Combinatorial 3-Manifolds

### 2013-06-10 paper/paper.tex
more writing. expanded abstract, fixed some figs

\begin{abstract}
  This paper presents a discrete version of Grove and Shiohama's generalized sphere theorem. TO DO: EXPAND ABSTRACT?
\end{abstract}

@@

\noindent Our results are derived from brute-force checking of a combinatorial 3-manifold census created by Lutz and Sullivan \cite{LS}.

@@

\begin{thm}[Trout] For any positively curved combinatorial 3-manifold $M$ we have $diam_1(M)\leq 5$. Moreover, this bound is sharp.
\label{thm:discrete_BM}
\end{thm}

@@

\begin{dfn}[Hops] Suppose $\tau$ is a 2-simplex in $M$ and $v_1$ and $v_2$ are vertices in $M$ such that $v_1*\tau$ and $v_2*\tau$ are 3-simplices in $M$. The PL-path from $v_1$ through the barycenter of $\tau$ and ending on $v_2$ will be called a \textbf{hop} from $v_1$ to $v_2$. See Figure \ref{fig:hop}. TO DO: FIX THIS FIG NUM  VADAS: I THINK THIS IS DUE TO AMSTHM AND I``M NOT SURE HOW TO FIX IT.
\end{dfn}

\begin{figure}
	\label{fig:hop}
    \begin{center}
        \includegraphics[width=0.6\linewidth]{figures/hops.pdf}
        \caption{A {\em hop} from vertex $v_1$ to vertex $v_2$. TO DO: CROP OUT LEFT IMAGE}
    \end{center}
\end{figure}

@@

\begin{dfn}[Jumps] Suppose there are edges $e_1$ and $e_2$ and vertices $v_1$ and $v_2$ in $M$ so that $e_1*e_2$ is a 3-simplex in $M$ and $v_1*e_1$ and $v_2*e_2$ are 2-simplices in $M$. We call the PL-path from $v_1$ through the barycenters of $e_1$ and $e_2$ and ending on $v_2$ a \textbf{jump} from $v_1$ to $v_2$. See Figure \ref{fig:jump}. TO DO: FIX THIS FIG NUM
\end{dfn}

\begin{figure}
    \label{fig:jump}
    \begin{center}
        \includegraphics[width=0.4\linewidth]{figures/jump.pdf}
        \caption{A {\em jump} from vertex $v_1$ to vertex $v_2$}
    \end{center}
\end{figure}

\noindent Just as for edges in an edge-path, the length of each hop and jump will be its length as a PL-path in the standard PL metric. Some Euclidean geometry tells us these lengths are \begin{equation} \label{eqn:jump_length} H = \frac{2}{3}\sqrt{2} \end{equation} and \begin{equation} \label{eqn:jump_length} J = \frac{1}{2}\sqrt{2} + \sqrt{3} \end{equation} respectively. We let $d(v,w)$ denote the \textbf{distance between vertices $v$ and $w$} obtained by minimizing over all paths containing edges as well as hops and jumps. Similarly, we let $diam(M)$ denote the \textbf{diameter of $M$} defined in terms of the distance function $d$. Let $\mathcal{P}$ denote the set of all the paths containing edges, hops or jumps.

@@

\begin{thm} If $M$ is a positively curved combinatorial 3-manifold then $diam(M) \leq 2J$.
\label{thm:discrete_BM_expanded_paths}
\end{thm}

@@

\begin{thm} Any positively curved combinatorial 3-manifold $M$ with $diam(M)>J$ is homeomorphic to the 3-sphere. This bound is sharp, and equal to half the maximum diameter which occurs for any such $M$.
\label{thm:discrete_GS}
\end{thm}

@@

Here is a table of the number of triangulations in the census at each topological type and diameter. \vspace{.1in}

\begin{tabular} {| c | c | c | c | c | c | c |}
    \hline
Diameter & $S^{3}/Q$ & $L(4,1)$ & $L(3,1)$ & $RP^{3}$ & $S^{3}$ & Total \\
\hline
\hline
$1E$&  &  &  &   &3    & 3    \\
$1H$&1 &  &  &10 &173  & 184  \\
$2E$&  &  &1 &7  &401  & 409  \\
$1J$&  &1 &1 &5  &2438 & 2445 \\
$3E$&  &  &  &   &1060 & 1060 \\
$2H$&  &  &  &   &582  & 582  \\
$4E$&  &  &  &   &93   & 93   \\
$2J$&  &  &  &   &11   & 11   \\
\hline
Total&1 &1 &2 &22   &4761   &4787  \\
\hline
\end{tabular}

\noindent Here is a table of the number of triangulations at each diameter.

\begin{figure}
    \begin{center}
    \includegraphics[width=0.9\linewidth]{figures/global_statistics.png}
    \caption{Global statistics. TO DO: CHANGE DIAMETER LABELS TO $1, H, 2, J, 3, 2H, 4, 2J$ AND PUT THESE ON HORIZONTAL AXIS. TO DO: CHANGE GRAPH TO SHOW SEPARATE DATA POINTS FOR EACH TOPOLOGICAL TYPE. TO DO: CREATE ANOTHER GRAPH WITH JUST EDGE DIAMETER COUNTS, WHERE THE CORRESPONDENCE BETWEEN DIAMETER(s) AND EDGE-DIAMETER IS $0\rightarrow 0, 1\rightarrow 1, \{H, 2\}\rightarrow 2, \{J,3\}\rightarrow 3, 4\rightarrow 4, 2J\rightarrow 5$}
    \end{center}
    \label{global_statistics}
\end{figure}

@@

This important property is restored if we include hops and jumps. In \cite{Trout10} the following is shown.

>>>>

\begin{abstract}
The generalized sphere theorem of Grove and Shiohama is a central result in the differential geometry of positively curved manifolds. It shows that if $M$ is a Riemannian manifold whose sectional curvature is everywhere greater than a positive constant and $M$ has diameter greater than half the maximum allowed by the Bonnet-Myers theorem, then $M$ must be homeomorphic to a sphere. In this paper, we present a novel discrete version of this theorem which applies to positively curved {\em combinatorial} 3-manifolds. That is, those for which at most five tetrahedra are incident at each edge. In the standard piecewise-linear metric, this bound is equivalent to requiring an angle deficit along each edge. Such spaces have been studied previously and found to satisfy a beautiful discrete version of the Bonnet-Myers theorem. Here we prove a corresponding {\em discrete} generalized sphere theorem using a complete census of positively curved 3-manifolds completed by Lutz and Sullivan.
\end{abstract}

@@

\noindent The main result of this paper is a discrete version of Theorem \ref{thm:grove_shiohama} which is proved by brute-force checking of a combinatorial 3-manifold census created by Lutz and Sullivan \cite{LS}. Before we can state precisely this result, we need a few preliminaries.

@@

\begin{thm}[Trout] For any connected positively curved combinatorial 3-manifold $M$ we have $diam_1(M)\leq 5$. Moreover, this bound is sharp.
\label{thm:discrete_BM}
\end{thm}

@@

\begin{dfn}[Hops] Suppose $\tau$ is a 2-simplex in $M$ and $v_1$ and $v_2$ are vertices in $M$ such that $v_1*\tau$ and $v_2*\tau$ are 3-simplices in $M$. The PL-path from $v_1$ through the barycenter of $\tau$ and ending on $v_2$ will be called a \textbf{hop} from $v_1$ to $v_2$. See Figure \ref{fig:hop} for an illustration.
\end{dfn}

\begin{figure}
	\centering
		\includegraphics[width=0.23\linewidth]{figures/hops.pdf}
    \caption{A {\em hop} from vertex $v_1$ to vertex $v_2$.}
    	\label{fig:hop}
\end{figure}

@@

\begin{dfn}[Jumps] Suppose there are edges $e_1$ and $e_2$ and vertices $v_1$ and $v_2$ in $M$ so that $e_1*e_2$ is a 3-simplex in $M$ and $v_1*e_1$ and $v_2*e_2$ are 2-simplices in $M$. We call the PL-path from $v_1$ through the barycenters of $e_1$ and $e_2$ and ending on $v_2$ a \textbf{jump} from $v_1$ to $v_2$. For an illustration, see Figure \ref{fig:jump}.
\end{dfn}

\begin{figure}
	\centering
		\includegraphics[width=0.30\linewidth]{figures/jump.pdf}
    \caption{A {\em jump} from vertex $v_1$ to vertex $v_2$}
    \label{fig:jump}
\end{figure}

Just as for edges in an edge-path, the length of each hop and jump will be its length as a PL-path in the standard PL metric. Some Euclidean geometry tells us these lengths are \begin{equation} \label{eqn:jump_length} H = \frac{2}{3}\sqrt{2} \end{equation} and \begin{equation} \label{eqn:jump_length} J = \frac{1}{2}\sqrt{2} + \sqrt{3} \end{equation} respectively. We let $d(v,w)$ denote the \textbf{distance between vertices $v$ and $w$} obtained by minimizing over all paths containing edges as well as hops and jumps. Similarly, we let $diam(M)$ denote the \textbf{diameter of $M$} defined in terms of the distance function $d$. Let $\mathcal{P}$ denote the set of all the paths containing edges, hops or jumps.

@@

\begin{thm}[Trout] For any positively curved combinatorial 3-manifold $M$ we have $diam(M) \leq 2J$.
\label{thm:discrete_BM_expanded_paths}
\end{thm}

@@

\begin{thm} Any connected positively curved combinatorial 3-manifold $M$ with $diam(M)>J$ is homeomorphic to the 3-sphere. This bound is sharp, and equal to half the maximum diameter which occurs for any such $M$.
\label{thm:discrete_GS}
\end{thm}

@@

Here is a table of the number of triangulations in the census at each topological type and diameter.

\begin{figure}
\centering
\begin{tabular} {| r | r | r | r | r | r | r |}
\hline
Diameter & $S^{3}/Q$ & $L(4,1)$ & $L(3,1)$ & $RP^{3}$ & $S^{3}$ & Total \\
\hline
\hline
$1$&  &  &  &   &3    & 3    \\
$H$&1 &  &  &10 &173  & 184  \\
$2$&  &  &1 &7  &401  & 409  \\
$J$&  &1 &1 &5  &2438 & 2445 \\
$3$&  &  &  &   &1060 & 1060 \\
$2H$&  &  &  &   &582  & 582  \\
$4$&  &  &  &   &93   & 93   \\
$2J$&  &  &  &   &11   & 11   \\
\hline
Total&1 &1 &2 &22   &4761   &4787  \\
\hline
\end{tabular}
\caption{The positively curved combinatorial 3-manifolds by diameter and topological type}
\label{fig:manifold_counts}
\end{figure}

\noindent Compare this table with the corresponding counts indexed by the coarser {\em edge-diameter}.

\begin{figure}
\centering
\begin{tabular} {| r | r | r | r | r | r | r |}
\hline
Diameter & $S^{3}/Q$ & $L(4,1)$ & $L(3,1)$ & $RP^{3}$ & $S^{3}$ & Total \\
\hline
\hline
$1$ &  &  &  &   &3    & 3    \\
$2$ & 1&  &1 &17 &574  & 593  \\
$3$ &  &1 &1 &5  &3498 & 3505 \\
$4$ &  &  &  &   &675  & 675   \\
$5$ &  &  &  &   &11   & 11   \\
\hline
Total&1 &1 &2 &22   &4761   &4787  \\
\hline
\end{tabular}
\caption{The positively curved combinatorial 3-manifolds by edge-diameter and topological type}
\label{fig:manifold_counts}
\end{figure}

\begin{figure}
    \begin{center}
    \includegraphics[width=0.6\linewidth]{figures/global_statistics.png}
    \caption{Global statistics. TO DO: CHANGE DIAMETER LABELS TO $1, H, 2, J, 3, 2H, 4, 2J$ AND PUT THESE ON HORIZONTAL AXIS. TO DO: CHANGE GRAPH TO SHOW SEPARATE DATA POINTS FOR EACH TOPOLOGICAL TYPE. TO DO: CREATE ANOTHER GRAPH WITH JUST EDGE DIAMETER COUNTS, WHERE THE CORRESPONDENCE BETWEEN DIAMETER(s) AND EDGE-DIAMETER IS $0\rightarrow 0, 1\rightarrow 1, \{H, 2\}\rightarrow 2, \{J,3\}\rightarrow 3, 4\rightarrow 4, 2J\rightarrow 5$}
    \end{center}
    \label{global_statistics}
\end{figure}

@@

This important property is restored, at least for positively curved manifolds, if we include hops and jumps. In \cite{Trout10} the following is shown.
